---
title: 'IMMERSE day 4: text wrangling basics'
author: "Casey O'Hara"
date: "2023-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r attach packages}
library(tidyverse)
library(here)
```

- intro to stringr
- intro to regex
- fix typos in state dataset and save back out

# Intro to `stringr` functions

Here we'll play a little with some basic `stringr` functions, and pattern vs. vector of strings.  Detecting strings (for filtering) and replacing/removing strings (for cleaning up text) are hugely helpful; other stringr functions certainly have their uses as well.  Here we're focusing on standardizing values in a character column so we can assign meaningful levels as a factor, to potentially use the column as a categorical variable.

- `str_match`, `str_match_all`
- `str_detect`
- `str_split`
- `str_replace`, `str_replace_all`
- `str_subset`, `str_count`, `str_locate`
- `str_trim`, `str_squish`
- `str_to_upper/lower/title/sentence`

Solicit pet names from the fellows to create a vector of character type; add in some random spaces and capitalization and punctuation

``` {r}
x <- c("Waffle!!!", "teddy", "  critter?!", "  taco   CAT   !")
stringr::str_to_title(x)
str_to_lower(x)
str_squish(x) %>% 
  str_to_upper()
str_squish(x) %>% 
  str_to_sentence()

str_split(x, 'e'); str_split(x, 't')

str_replace(x, 'e', 'Z')
str_replace(x, ' ', '_')
str_replace_all(x, ' ', '_')
str_replace_all(x, ' +', '_') ### replace one or more spaces with _
str_replace_all(str_squish(x), ' +|!+|\\?+', '_')

str_detect(x, 't'); str_detect(x, 'dd|tt') ### is pattern in the string?
```
## OK now let's put into practice!  You do:

Read in the file of `ca_schools.csv` we created in the previous script. Fix the `county` column to some standard format (however you prefer).  Note there is a typo in some instances of the Santa Barbara county name; identify those and fix those.  

When finished, check the unique values found in the county column; there should be 45 unique values, no more, no less.  Once you've figured out a standard format that works well, then write out as `ca_schools_clean.csv`.

```{r}
ca_schools <- read_csv(here('data/ca_schools.csv'))

# ca_schools$county %>% unique()

ca_schools_clean <- ca_schools %>%
  mutate(county = str_replace_all(county, '_', ' ')) %>%
  mutate(county = str_to_title(county)) %>%
  mutate(county = str_replace(county, 'Barabara', 'Barbara')) %>%
  mutate(county = factor(county))

ca_schools_clean$county %>% unique()

write_csv(ca_schools_clean, here('data/ca_schools_clean.csv'))

```

## Wrap up - any other text wrangling needs/questions?

